---
title: "Untitled"
author: "Liuyixin Shao and Yi Su"
date: "2023-11-09"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(scales)
library(usmap)
```

```{r}
df <- read_csv('Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System.csv')
```

# Some random overview of the dataset

```{r}
colnames(df)
```

```{r}
unique(df%>% select('Topic'))
```

```{r}
unique(df%>% select('Race/Ethnicity'))
```

```{r}
unique(df%>% select('Total'))
```

```{r}
df %>% distinct(Income)
```

```{r}
range(df$YearEnd)
```

Analyzing obesity rates by race is critical to understanding differences
in health outcomes across racial and ethnic groups. Such analyses can
help to identify the underlying sociocultural, economic, and
environmental factors that contribute to varying obesity prevalence
rates, thereby drawing attention to addressing health inequities and
scaling up more effective public health initiatives within a given
group. Here, the question we want to address is "Do obesity rates vary
by race?" More specifically, we would like to explore this issue in
terms of geographical location and year. We decided to use the
"Nutrition, Physical Activity, and Obesity - Behavioral Risk Factor
Surveillance System" dataset that we found on data.gov. The columns we
choose to use to solve this question in the dataset are "Race.Ethnicity"
(for race), "Data_Value_Alt" (for the BMI), "Geolocation" (for the
geographical location), and "Year End" (for year). We will filter out
the dataset by "Class == Obesity / Weight Status" since we only care
about the obesity rate in this dataset. Since the dataset contains the
necessary information, it is possible to answer the questions with this
dataset.

# Plot of obesity rate by race (national/ all year)

The proportion of adults with a body mass index (BMI) equal to or
greater than 30 ($\geq 30 kg/m^2$) based on self-reported weight and height

2 notes:

The obesity rate is defined as the proportion of adults with a body mass
index (BMI) $\geq 30 kg/m^2$

Abbreviations comparison table:

"AS" = "Asian" 

"OTR" = "Other" 

"2+" = "2 or more races" 

"NHW" = "Non-Hispanic White"

"HPI" = "Hawaiian/Pacific Islander" 

"AI/AN" = "American Indian/Alaska Native" 

"HISP" = "Hispanic" 

"NHB" = "Non-Hispanic Black"

```{r}
df_race <- df %>%
  filter(LocationAbbr == 'US') %>%
  filter(Class == 'Obesity / Weight Status') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  select(`Race/Ethnicity`, Data_Value_Alt) %>%
  na.omit()
```

```{r}
df_summary_race <- df_race %>%
  group_by(`Race/Ethnicity`) %>%
  summarise(Avg_Data_Value_Alt = mean(Data_Value_Alt))

df_summary_race %>% arrange(Avg_Data_Value_Alt)

df_summary_race$`Race/Ethnicity` <- factor(df_summary_race$`Race/Ethnicity`, levels = c("Asian", "Other", "Non-Hispanic White", "2 or more races", "Hispanic", "Hawaiian/Pacific Islander", "American Indian/Alaska Native", "Non-Hispanic Black"))
```

```{r, fig.width= 5, fig.height= 4}
average_obesity_rate <- mean(df_race$Data_Value_Alt, na.rm = TRUE)
p1 <- ggplot(df_summary_race, aes(x=`Race/Ethnicity`, y=Avg_Data_Value_Alt)) +
  geom_bar(stat="identity", fill = "#404788FF") + 
  labs(y= "The Obseity Rate",
       x = "Race / Ethnicity", 
       title = "National Average Obesity Rates by Race, 2011-2021",
       caption = "Source: Centers for Disease Control and Prevention") + 
  scale_y_continuous(limits=c(0, 40),
                     breaks=seq(0, 40, 5),
                     labels = c("0%", "5%", "10%", "15%", "20%", "25%", "30%", "35%", "40%")) + 
  scale_x_discrete(
    labels = c(
      "Asian" = "AS",
      "Other" = "OTR", 
      "Non-Hispanic White" = "NHW",
      "2 or more races" = "2+",
      "Hispanic" = "HISP",
      "Hawaiian/Pacific Islander" = "HPI",
      "American Indian/Alaska Native" = "AI/AN",
      "Non-Hispanic Black" = "NHB"
      )
    ) +
  theme_light() + 
  theme(axis.text.x=element_text(color = "black"),
        axis.text.y=element_text(color = "black"),
        panel.grid.major=element_line(color="Gray", linetype=1),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),         
        plot.title = element_text(size = 12.5, face = 'bold'),
        plot.caption.position = 'plot',
        plot.caption = element_text(size = 7.5)) +
  geom_hline(yintercept = mean(df_race$Data_Value_Alt), linetype = "dashed", color = "red", size = 1) +
  geom_text(aes(x = 2, y = average_obesity_rate, label = "Average Obesity Rate"), 
            color = "red", 
            size = 3, 
            vjust = -1)

p1 
```
```{r}
ggsave("451_final_image_1.png", p1, width = 5, height = 4, dpi = 300)
```

We selected a bar plot to represent the average obesity rates across different races nationwide from 2011 to 2021. With Diagonalization, overall, we can see that the ranking of the average nationwide obesity rate for each race, from lowest to highest, is "Asian," "Other," "2 or more races," "Non-Hispanic White," "Hawaiian/Pacific Islander," "American Indian/Alaska Native," "Hispanic," and "Non-Hispanic Black." More specifically, the graph shows that Asians have the lowest average nationwide obesity rate (below 15%) relative to other racial groups, with a considerable gap. All other racial categories show national average obesity rates exceeding 25%, with other races and non-Hispanic whites in the 25-30 percent range; 2 or more races, Hispanic, and Native Hawaiian/Pacific Islander in the 30-35 percent range; and American Indian/Alaska Native and non-Hispanic blacks in the 35-40 percent range. Of these, the Non-Hispanic Black population registering the highest average rate. We also added a red dashed line to represent the average obesity rate over all races. We know from the graph that the average nationwide obesity rate for all race is very close to 30% (the true average nationwide obesity rate stands at 29.92%). As illustrated by the plot, with the exception of Asians, Others, and Non-Hispanic Whites, all other groups exceed this average.

option: national/ state/ year/ income? (possible)

# Plot of obesity rate by state (all year/ all race)

```{r}
df_state <- df %>%
  filter(Class == 'Obesity / Weight Status') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  filter(LocationAbbr != "US") %>%
  select(LocationAbbr, Data_Value_Alt) %>%
  na.omit()
```

```{r}
df_summary_state <- df_state %>%
  group_by(LocationAbbr) %>%
  summarise(Avg_Data_Value_Alt = mean(Data_Value_Alt))

midpoint <- median(df_summary_state$Avg_Data_Value_Alt)

df_summary_state  <- df_summary_state  %>%
  rename(state = LocationAbbr)
```

```{r}
plot_usmap(data = df_summary_state, values = "Avg_Data_Value_Alt", default_fill = "gray") + 
  scale_fill_gradient2(low = "green",
                       mid = "yellow",
                       high = "red", 
                       midpoint = midpoint,
                       name = "legend") +
  labs(title = "Map of Obesity Rate by State",
       caption = "Source: Centers for Disease Control and Prevention") +
  theme(plot.title = element_text(size = 19, hjust = 0.5),
        plot.title.position = "plot",
        legend.title = element_text(size = 13, hjust = 0.5),
        legend.text = element_text(size = 10, hjust = 0.5))
```

option: year/ race/ ?income

# Plot of obesity rate by year (national/ all race)
```{r}
df_year <- df %>% 
  filter(Class == 'Obesity / Weight Status') %>% 
  filter(LocationAbbr == 'US') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  select(YearEnd, Data_Value_Alt, `Race/Ethnicity`) %>%
  na.omit
#df_year %>% distinct(Question)
```

```{r}
last_values <- df_year %>%
  group_by(`Race/Ethnicity`) %>%
  summarize(last_value = last(Data_Value_Alt)) %>%
  arrange(desc(last_value))

df_year$`Race/Ethnicity` <- factor(df_year$`Race/Ethnicity`, 
                                   levels = last_values$`Race/Ethnicity`)

custom_labels <- c(
      "Asian" = "AS",
      "Other" = "OTR", 
      "Non-Hispanic White" = "NHW",
      "2 or more races" = "2+",
      "Hispanic" = "HISP",
      "Hawaiian/Pacific Islander" = "HPI",
      "American Indian/Alaska Native" = "AI/AN",
      "Non-Hispanic Black" = "NHB"
      )

ggplot(df_year, aes(x = YearEnd, y = Data_Value_Alt, color = `Race/Ethnicity`)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Trend of Obesity Rate by Race/Ethnicity (2011-2021)",
       x = "Year",
       y = "Obesity Rate (%)",
       caption = "Source: Centers for Disease Control and Prevention") +
  scale_color_viridis_d(labels = custom_labels) +
  theme(plot.title = element_text(size = 19),
        axis.text.y = element_text(size =13),
        axis.text.x = element_text(size =13),
        axis.title.y = element_text(size =15),
        axis.title.x = element_text(size =15),
        legend.title = element_text(size =15),
        legend.text = element_text(size =13),) +
  scale_x_continuous(name = "Year", breaks = seq(2010, 2021, by = 2)) +
  scale_y_continuous(labels = percent_format(scale = 1))
```
