---
title: "451 Group Project"
author: "Liuyixin Shao and Yi Su"
date: "2023-11-09"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(scales)
library(usmap)
```

```{r}
df <- read_csv('Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System.csv')
```

# Some random overview of the dataset

```{r}
colnames(df)
```

```{r}
unique(df%>% select('Topic'))
```

```{r}
unique(df%>% select('Race/Ethnicity'))
```

```{r}
unique(df%>% select('Total'))
```

```{r}
df %>% distinct(Income)
```

```{r}
range(df$YearEnd)
```

Analyzing obesity rates by race is critical to understanding differences
in health outcomes across racial and ethnic groups. Such analyses can
help to identify the underlying sociocultural, economic, and
environmental factors that contribute to varying obesity prevalence
rates, thereby drawing attention to addressing health inequities and
scaling up more effective public health initiatives within a given
group. Here, the question we want to address is "Do obesity rates vary
by race?" More specifically, we would like to explore this issue in
terms of geographical location and year. We decided to use the
"Nutrition, Physical Activity, and Obesity - Behavioral Risk Factor
Surveillance System" dataset that we found on data.gov. The columns we
choose to use to solve this question in the dataset are "Race.Ethnicity"
(for race), "Data_Value_Alt" (for the BMI), "Geolocation" (for the
geographical location), and "Year End" (for year). We will filter out
the dataset by "Class == Obesity / Weight Status" since we only care
about the obesity rate in this dataset. Since the dataset contains the
necessary information, it is possible to answer the questions with this
dataset.

# Plot of obesity rate by race (national/ all year)

The proportion of adults with a body mass index (BMI) equal to or
greater than 30 ($\geq 30 kg/m^2$) based on self-reported weight and height

2 notes:

The obesity rate is defined as the proportion of adults with a body mass
index (BMI) $\geq 30 kg/m^2$

Abbreviations comparison table:

"AS" = "Asian" 

"OTR" = "Other" 

"2+" = "2 or more races" 

"NHW" = "Non-Hispanic White"

"HPI" = "Hawaiian/Pacific Islander" 

"AI/AN" = "American Indian/Alaska Native" 

"HISP" = "Hispanic" 

"NHB" = "Non-Hispanic Black"

```{r}
df_race <- df %>%
  filter(LocationAbbr == 'US') %>%
  filter(Class == 'Obesity / Weight Status') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  select(`Race/Ethnicity`, Data_Value_Alt) %>%
  na.omit()
```

```{r}
df_summary_race <- df_race %>%
  group_by(`Race/Ethnicity`) %>%
  summarise(Avg_Data_Value_Alt = mean(Data_Value_Alt))

df_summary_race %>% arrange(Avg_Data_Value_Alt)

df_summary_race$`Race/Ethnicity` <- factor(df_summary_race$`Race/Ethnicity`, levels = c("Asian", "Other", "Non-Hispanic White", "2 or more races", "Hispanic", "Hawaiian/Pacific Islander", "American Indian/Alaska Native", "Non-Hispanic Black"))
```

```{r, fig.width= 5, fig.height= 4}
average_obesity_rate <- mean(df_race$Data_Value_Alt, na.rm = TRUE)
p1 <- ggplot(df_summary_race, aes(x=`Race/Ethnicity`, y=Avg_Data_Value_Alt)) +
  geom_bar(stat="identity", fill = "#404788FF") + 
  labs(y= "The Obseity Rate",
       x = "Race / Ethnicity", 
       title = "National Average Obesity Rates by Race, 2011-2021",
       caption = "Source: Centers for Disease Control and Prevention") + 
  scale_y_continuous(limits=c(0, 40),
                     breaks=seq(0, 40, 5),
                     labels = c("0%", "5%", "10%", "15%", "20%", "25%", "30%", "35%", "40%")) + 
  scale_x_discrete(
    labels = c(
      "Asian" = "AS",
      "Other" = "OTR", 
      "Non-Hispanic White" = "NHW",
      "2 or more races" = "2+",
      "Hispanic" = "HISP",
      "Hawaiian/Pacific Islander" = "HPI",
      "American Indian/Alaska Native" = "AI/AN",
      "Non-Hispanic Black" = "NHB"
      )
    ) +
  theme_light() + 
  theme(axis.text.x=element_text(color = "black"),
        axis.text.y=element_text(color = "black"),
        panel.grid.major=element_line(color="Gray", linetype=1),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),         
        plot.title = element_text(size = 12.5, face = 'bold'),
        plot.caption.position = 'plot',
        plot.caption = element_text(size = 7.5)) +
  geom_hline(yintercept = mean(df_race$Data_Value_Alt), linetype = "dashed", color = "red", size = 1) +
  geom_text(aes(x = 2, y = average_obesity_rate, label = "Average Obesity Rate"), 
            color = "red", 
            size = 3, 
            vjust = -1)

p1 
```
```{r}
ggsave("451_final_image_1.png", p1, width = 5, height = 4, dpi = 300)
```

option: national/ state/ year/ income? (possible)

### Analysis for Plot 1: National Average Obesity Rates by Race, 2011-2021

We selected a bar plot to represent the average obesity rates across different races nationwide from 2011 to 2021. From this bar plot, we can see that the national average obesity rate does vary by race between 2011 and 2021. With Diagonalization, overall, we see that the ranking of the average nationwide obesity rate for each race, from lowest to highest, is "Asian," "Other," "2 or more races," "Non-Hispanic White," "Hawaiian/Pacific Islander," "American Indian/Alaska Native," "Hispanic," and "Non-Hispanic Black." More specifically, the graph shows that Asians have the lowest average nationwide obesity rate (below 15%) relative to other racial groups, with a considerable gap. All other racial categories show national average obesity rates exceeding 25%, with other races and non-Hispanic whites in the 25-30 percent range; 2 or more races, Hispanic, and Native Hawaiian/Pacific Islander in the 30-35 percent range; and American Indian/Alaska Native and non-Hispanic blacks in the 35-40 percent range. Of these, the Non-Hispanic Black population registering the highest average rate. We also added a red dashed line to represent the average obesity rate over all races. We know from the graph that the average nationwide obesity rate for all race is very close to 30% (the true average nationwide obesity rate stands at 29.92%). As illustrated by the plot, with the exception of Asians, Others, and Non-Hispanic Whites, all other groups exceed this average.


# Plot of obesity rate by state (all year/ all race)

```{r}
df_state <- df %>%
  filter(Class == 'Obesity / Weight Status') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  filter(LocationAbbr != "US") %>%
  select(LocationAbbr, Data_Value_Alt) %>%
  na.omit()
```

```{r}
df_summary_state <- df_state %>%
  group_by(LocationAbbr) %>%
  summarise(Avg_Data_Value_Alt = mean(Data_Value_Alt))

# midpoint <- median(df_summary_state$Avg_Data_Value_Alt)

df_summary_state  <- df_summary_state  %>%
  rename(state = LocationAbbr)
```

```{r, fig.width= 7.3, fig.height= 5}
p2 <- plot_usmap(data = df_summary_state, values = "Avg_Data_Value_Alt", default_fill = "gray") + 
  scale_fill_gradient(low = "white",
                      high = muted("#404788FF"), 
                      name = "Avg Obs Rate",
                      labels = scales::percent_format(scale = 1)) +
  labs(title = "Map of Average Obesity Rate by State, 2011-2021",
       caption = "Source: Centers for Disease Control and Prevention") +
  theme(plot.title = element_text(size = 19, hjust = 0.5, face = "bold"),
        plot.title.position = "plot",
        legend.title = element_text(size = 12, hjust = 0.5),
        legend.text = element_text(size = 10, hjust = 0.5),
        legend.position = c(0.9, 0.1),
        plot.caption.position = 'plot',
        plot.caption = element_text(size = 10),
        plot.background = element_rect(fill = "white", color = "white"))
p2
```

```{r}
ggsave("451_final_image_2.png", p2, width = 7.3, height = 5, dpi = 300)
```

option: year/ race/ ?income

### Analysis for Plot 2: Map of Average Obesity Rate by State, 2011-2021
We used a map to show the average obesity rate for all races in each state from 2011 to 2021. To be color-blind friendly, here we've used only one color in the legend and used the shade of the color to indicate the state's average obesity rate. The darker the color, the higher the obesity rate in the state. Since using color saturation as a comparison method leads to poor visual cues, for this map, we want our audience to read the approximate distribution of obesity rates by state, not specific values. Based on the categorization of U.S. regions from the National Geographic (see this link: https://education.nationalgeographic.org/resource/united-states-regions/), we can clearly see that the average obesity rate in the western United States is relatively low. Of these, Colorado has the lowest average obesity rate. The Northeast also has a relatively low average obesity rate. The Southeast, Midwest, and Southwest regions have the highest average obesity rates, with the exceptions of Arizona, New Mexico, Minnesota, Illinois, and Florida. At the same time, we can see that the two states with the highest obesity rates are West Virginia and Mississippi. 

(Note: We are well aware that using color saturation as a comparison method leads to poor visual cues. As a result, the shiny app we end up presenting will be multiple charts to improve the visual cues. In the shiny app, viewers can filter race to see the obesity rates of different races in each state. We would also like to make it possible for viewers to hover over a state on the map and have the corresponding state name and specific obesity rate displayed.)

# Plot of obesity rate by year (national/ all race)
```{r}
df_year <- df %>% 
  filter(Class == 'Obesity / Weight Status') %>% 
  filter(LocationAbbr == 'US') %>%
  filter(Question == 'Percent of adults aged 18 years and older who have obesity') %>%
  select(YearEnd, Data_Value_Alt, `Race/Ethnicity`) %>%
  na.omit
#df_year %>% distinct(Question)
```

```{r}
last_values <- df_year %>%
  group_by(`Race/Ethnicity`) %>%
  summarize(last_value = last(Data_Value_Alt)) %>%
  arrange(desc(last_value))

df_year$`Race/Ethnicity` <- factor(df_year$`Race/Ethnicity`, 
                                   levels = last_values$`Race/Ethnicity`)

custom_labels <- c(
      "Asian" = "AS",
      "Other" = "OTR", 
      "Non-Hispanic White" = "NHW",
      "2 or more races" = "2+",
      "Hispanic" = "HISP",
      "Hawaiian/Pacific Islander" = "HPI",
      "American Indian/Alaska Native" = "AI/AN",
      "Non-Hispanic Black" = "NHB"
      )
```
```{r, fig.width= 7, fig.height= 6}
p3<- ggplot(df_year, aes(x = YearEnd, y = Data_Value_Alt, color = `Race/Ethnicity`)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "Trend of Obesity Rate by Race/Ethnicity, 2011-2021",
       x = "Year",
       y = "The Obesity Rate",
       caption = "Source: Centers for Disease Control and Prevention") +
  scale_color_discrete(labels = custom_labels) +
  theme(plot.title = element_text(size = 19, hjust = 0.2, face = "bold"),
        axis.text.y = element_text(size =13),
        axis.text.x = element_text(size =13),
        axis.title.y = element_text(size =15),
        axis.title.x = element_text(size =15),
        legend.title = element_text(size =15),
        legend.text = element_text(size =13),
        plot.caption.position = 'plot',
        plot.caption = element_text(size = 10),
        plot.background = element_rect(fill = "white", color = "white")) +
  scale_x_continuous(name = "Year", breaks = seq(2010, 2021, by = 2)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, NA))
p3
```

```{r}
ggsave("451_final_image_3.png", p3, width = 7, height = 6, dpi = 300)
```


### Analysis for Plot 3: Trend of Obesity Rate by Race/Ethnicity, 2011-2021
We used a line plot to show the trend in obesity rates between 2011 and 2021. Different colors of lines represent different races, such as Asian, Non-Hispanic White, Non-Hispanic Black, Hispanic, Hawaiian/Pacific Islander, American Indian/Alaska Native, 2 or more races, and Others. The legend on the right side of the figure helps in interpreting each line with a different color. The x-axis indicates the year (2011-2021) and the y-axis indicates the obesity rate. 
Looking at the plot, we can see that obesity rates for all races have a slight upward trend from 2011 to 2021. In terms of trends, we can see that Asian, Non-Hispanic White, Hispanic, and Non-Hispanic Black obesity rates are increasing at a relatively steady rate. Trends in obesity rates for other races, 2 or more races, American Indian/Alaska Native, and Hawaiian/Pacific Islander are growing but unstable (with many ups and downs)). Comparing the obesity rates of all races, the obesity rate of Asians has been around 10% which is significantly lower than other races. All other races have obesity rates of 20% or more. Meanwhile, we can see that the Non-Hispanic Blacks have a consistently high obesity rate, while Hawaii/Pacific Islanders have the fastest growing obesity rate (along with the largest upward and downward fluctuations), increasing from about 25 percent in 2011 to more than 40 percent by 2021. The increasing obesity rate year after year warrants more attention to physical health as well as obesity management.
